### 2012-07-18 Doc/VRIPHYS2012/vriphys2012.tex
more on the article
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@425 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@

Objects with thin structure form a very specific field of research. Because
they have high ratio between width and thickness methods from volumetric
deformation modeling cannot be immediately used. Such object occur in many
places e.g. textiles, paper or leaves. They are also very common in
anatomical structures for example in form of tubular objects like blood
vessels or colon, in the form of thin membranes, but also many others
(lens, skin, \ldots).

Methods based on continuum mechanics and namely theory of elasticity gained
on popularity ever since Terzopoulos \cite{Terzopoulos1987} presented his
work on elastic deformation modeling. While the methods using finite
elements and theory of elasticity are fairly popular in volumetric modeling
\footnote{TODO: do we need ref. here?} of deformations, they are still
frowned upon in the area of thin structures. Instead either the
mass-spring models \cite{Volino2009} or bending models
\cite{Grinspun2003,Choi2007} are preferred. Unfortunately they have the
disadvantage that physical parameters of a material (like Young's modulus
and Poisson's ratio) cannot be used directly.

The usual argument against FEM based methods being that they are too slow.
This is indeed true if lots of elements are used to discretize the object.
However, exploiting the bending property of the shell elements allows us to
achieve good results even with relatively small number of elements. We can
use high-polygonal mesh only for rendering or collision detection and
response. Similar idea has been already presented in the work of Bouthors
et al. \cite{Bouthors2007}. They used twinned meshes for animation of mesh
defined by implicit surface. Simple low-polygonal mesh was used for
mechanical FEM model and the mesh was subdivided in areas with high
curvature after every step of the simulation to produce finer mesh for
rendering. The underlaying FEM model is however not aware of the implicit
surface used. We build upon the work of Comas \cite{Comas2010c} who used a
simple cubic polynomial to describe the element geometry and derived the
shell model from it.

The Comas's element has several drawbacks that we needed to address. Namely
it produces discontinuities between the elements on the edges, it is
missing one rotational degree of freedom and is not completely symmetric in
context of deformations.

% TODO: cite Ubach & Oñate
TODO: cite \cite{Ubach2010}: rotation-free shell elements, also based on
Bézier triangle

In this paper we present a new shell element based on the formulation of
Bézier triangle.

@@

% TODO
The main contributions of the paper are following:
\begin{itemize}
    % not really novel, similar model already in Ubach2010
    %\item shell element based on Bézier triangles
    \item Our element solves the problems of the element presented by Comas
        \cite{Comas2010c}.

    % TODO: needs better explanation
    % + two stage interpolation => Nodes (6DOFS) => bezier points => bezier
    %   triangle
    \item Two stage interpolation: 6DOF nodes -> bezier points -> bezier
        triangle

    % + separation between nodes and dofs
    \item The model separates nodes of the mesh from degrees of freedom of the
        control mesh. This allows mapping a surface on deformable skeleton.

    % + mapping: smooth surface results with less shells + collision response
    %   on smooth surface +  [? if I have time]  deformable on deformable:
    %   conjonctiva 2D motion on the surface of the shells
    \item Smooth maping of high resolution mesh on the curved surface of the
        shell. The mapped surfce can be used for collision detection and
        allows the propagation of proper collision response back onto the
        (low resolution) mechanical mesh.

@@

Shell element combines two types of deformations:

@@

    \item Elastic membrane: deformation in the plane of the element,
        described by in-plane displacements $u_x, u_y$ and rotation
        $\theta_z$ around the axis perpendicular to the plane of the
        element (Figure \ref{fig-membrane}).

    \item Bending plate: out of the plane deformation described by rotation
        around two in-plane axes $\theta_x, \theta_z$ and out of plane
        displacmenent $u_z$ (Figure \ref{fig-plate}).

@@

@@

The general $n$-th order Bézier triangle is defined as:

@@

Specificaly the cubic Bézier triangle ($n=3$) used in our model is
described by 10 control points and the surface is explicitly defined as (see
fig. \ref{fig-bezier}):

\begin{equation}\label{eq-cubicbez}
  \begin{split}
  T(\xi_1,\xi_2) & =
           + \xi_1^3\ P_1
           + \xi_2^3\ P_2
           + \xi_3^3\ P_3 \\
         & + 3\ \xi_1^2 \xi_2\ P_4
           + 3\ \xi_1^2 \xi_3\ P_5
           + 3\ \xi_2^2 \xi_3\ P_6 \\
         & + 3\ \xi_1 \xi_2^2\ P_7
           + 3\ \xi_1 \xi_3^2\ P_8
           + 3\ \xi_2 \xi_3^2\ P_9 \\
         & + 6\ \xi_1 \xi_2 \xi_3\ P_{10} \\
  \end{split}
\end{equation}

\noindent
where $ \xi_3 = 1 - \xi_1 - \xi_2 $. By naming the respective values of
the Bernstein basis functions $n_i$ we can shortly express
\eqref{eq-cubicbez} as:

@@

\Figure{0.8\linewidth}{bezier}
{Cubic Bézier triangle with net of 10 control points.}

@@

The Bézier triangles are defined by a net of control points, however the
simulated object is usually described by a triangular mesh. We construct
the Bézier mesh by taking into account the curvature of the triangular
mesh. When constructing the mesh a special care has to be taken to maintain
the continuity accross the nodes and edges. During the initialization phase
we partialy employ the method described in \cite{Ubach2010} to maintain
$C^0$ continuity on the edges. Each of the control points on the edge is
computed as the intersection of:

\begin{enumerate}
    \item The plane perpendicular to the normal at the vertex.
    \item The plane that contains the curve of triangle's contour. The
        choise is arbitrary, but necessary to maintain $C^0$ continuity. 
        We choose the plane defined by the edge of the flat triangle and
        average of the two normals at vertices of the edge. 
    \item The plane perpendicular to the edge of the flat triangle placed at
        $1/3$ of the edge length.
\end{enumerate}

@@

That way the central point is slightly elevated and not in the plane of
other nodes thus keeping the curvature of the element.

@@

In figure \ref{fig-fps} we present performance of the element in terms of
frames per second. All tests were performed on machine equipped Dual-Core
AMD Opteron Processor 2218 and 3 GB of RAM and conjugate gradient solver
has been to solve the system. Our implementation is without optimizations
and isn't parallel, which means only single CPU core was used for the
simulation. The values reported are for the raw physical simulation with
and without the frame fixing method described in subsection
\ref{sec-corot}. We also present results including the mapping of
high-resolution mesh onto the physical mesh for visualisation purposes or
collisions.

\Figure{\linewidth}{fps}
{Performance in frames per seconds for different number of elements. Shows
values for element with frame fixing (B) method described in subsection
\ref{sec-corot} and without it (A), and with high-resolution mesh of 2048
(C) or 3200 (D) triangles mapped onto the physical mesh.}

@@

@@

The element has been successfully tested in SOFA framework for surgical
simulations. The applications being for example simulation pre-operative
planning of cardiac surgery on arterial vessels \cite{Kislinskiy2012}, or
{\em something, something} of vagina/uterus.

% = = = = = = = = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
%
% TODO
% Other Applications:
% + cataract surgery
%

@@

On the other hand the choices provide a good trade-off between continuity
and simplicity. Only $ 3 \times 6~\mathrm{DOFs} = 18~\mathrm{DOFs}$ are
necessary while the Bézier triangle has theoretically $ 10 \times
3~\mathrm{DOFs} = 30~\mathrm{DOFs}$.

@@

    \item It is continuous on the edges and provides $C^1$ continuity in
    corners of the element. This is essential for good visual experience if
    high-resolution mesh is mapped on the physical mesh.

@@

Because of the formulation of Bézier triangle that is intrinsically present
there is a clean and obvious way of mapping high-resolution mesh for
better visualisation. We also provide a mapping of forces and velocities
between simulation mesh and physical mesh for collision detection/response
and interaction with the mesh which are essential component for the surgical
simulator.

>>>>

% For large/wide figures
%\FigureW{width}{name}{caption}
\newcommand{\FigureW}[3]{%
\begin{figure*}[htb]
  \centering
  \includegraphics[width=#1]{#2}
  \caption{\label{fig-#2}#3}
\end{figure*}}

@@

Objects with thin structure occur in many forms e.g. textiles, paper,
leaves etc. Such objects are also very common in anatomical structures for
example in the form of tubular objects like blood vessels or colon, in the
form of thin membranes, but also many others (lens, skin, \ldots). Because
they have high ratio between width and thickness methods for volumetric
deformation modeling cannot be immediately used. For this reason they form
a very specific field of research.

Methods based on continuum mechanics and namely theory of elasticity gained
on popularity in computer graphics and real-time simulators ever since
Terzopoulos \cite{Terzopoulos1987} presented his work on elastic
deformation modeling. While the methods using finite elements and theory of
elasticity are fairly popular in volumetric modeling \footnote{TODO: do we
need ref. here?} of deformations, they are still frowned upon in the area
of thin structures. Instead either the mass-spring models \cite{Volino2009}
or bending models \cite{Grinspun2003,Choi2007} are preferred. Unfortunately
they have the disadvantage that physical parameters of a material (like
Young's modulus and Poisson's ratio) cannot be used directly.

The usual argument against FEM based methods is that they are
computationally too expensive. This is indeed true if lots of elements are
used to discretize the object or non-linear models are used. However,
exploiting the bending property of the shell elements allows us to achieve
good results even with relatively small number of elements. We can use
high-polygonal mesh only for rendering or collision detection and response.
Similar idea has been already presented in the work of Bouthors et al.
\cite{Bouthors2007}. They used twinned meshes for animation of mesh defined
by implicit surface. Simple low-polygonal mesh was used for mechanical FEM
model and the mesh was subdivided in areas with high curvature after every
step of the simulation to produce finer mesh for rendering. The underlaying
FEM model is however not aware of the implicit surface used. We build upon
the work of Comas \cite{Comas2010c} who used a simple cubic polynomial to
describe the element geometry and derived the shell model from it. Comas's
element has several drawbacks that we needed to address. Namely it produces
discontinuities between the elements on the edges, it is missing one
rotational degree of freedom and is not completely symmetric in context of
deformations.

% TODO: integrate better to the rest of the section
Ubach and Oñate \cite{Ubach2010} also present a shell element based on
Bézier triangles. The element is however free from rotational degrees of
freedom which is unsuitable for physical simulators. It also reduces the
sparsity of stiffness matrix because of increased interconnection between
nodes.

% In this paper we present a new shell element based on the
% formulation of Bézier triangle.
%
% TODO: paper structure
TODO: paper structure

@@

% TODO
The main contributions of the paper are following:
\begin{itemize}
    % not really novel, similar model already in Ubach2010
    %\item shell element based on Bézier triangles
    \item Our element solves the problems of the element presented by Comas
        \cite{Comas2010c}: it is continuous on the element boundary and has
        all rotational degrees of freedom.

    % TODO: needs better explanation
    \item Two stage interpolation: 6DOF nodes -> bezier points -> bezier
        triangle

    % + separation between nodes and dofs
    %\item The model separates nodes of the mesh from degrees of freedom of the
    %    control mesh. This allows mapping a surface on deformable skeleton.

    % + mapping: smooth surface results with less shells + collision response
    %   on smooth surface 
    \item Mapping of high resolution mesh on the curved surface of the
        shell. The smooth mapped surface can be used for collision
        detection and allows the propagation of proper collision response
        back onto the (low resolution) mechanical mesh.

@@

Shell element is a combination of two simpler elements describing two types
of deformations:

@@

    \item Elastic membrane defining deformations in plane of the
    element. It encompasses deformations like stretching and shearing and
    is described by 3 degrees of freedom: in-plane displacements $u_x, u_y$
    and rotation $\theta_z$ around the axis perpendicular to the plane of
    the element (Figure \ref{fig-membrane}).

    \item Bending plate defining out of plane bending deformation. It is
    described by 3 degrees of freedom: rotation around two axes $\theta_x,
    \theta_z$ and out of plane displacement $u_z$ (Figure \ref{fig-plate}).

@@

Combining both elements we make use of all 6 degrees of freedom available
at each node of the element.

@@

Bézier triangles were designed in sixties by French mathematician de
Casteljau. As with other Bézier or B-spline surfaces the Bézier triangle is
defined by a mesh of control points that do not necessarily lie on the
surface. The general $n$-th order Bézier triangle requires $(n + 1)(n +
2)/2$ control points and it's surface is defined as:

@@

In our work we use cubic Bézier triangle ($n=3$). Triangle of smaller order
would fail to describe all possible deformations of the element and
triangles of higher order cannot be described unambiguously with only 18
DOFs available to us without adding further constraints on the control
mesh. Cubic triangle is described by 10 control points and the surface is
explicitly defined as (see fig. \ref{fig-bezier}):

\begin{equation}\label{eq-cubicbez}
  \begin{split}
  T(\xi_1,\xi_2) & =
             \xi_1^3\ P_1
           + \xi_2^3\ P_2
           + \xi_3^3\ P_3 \\
         & + 3\ \xi_1^2 \xi_2\ P_4
           + 3\ \xi_1^2 \xi_3\ P_5
           + 3\ \xi_2^2 \xi_3\ P_6 \\
         & + 3\ \xi_1 \xi_2^2\ P_7
           + 3\ \xi_1 \xi_3^2\ P_8
           + 3\ \xi_2 \xi_3^2\ P_9 \\
         & + 6\ \xi_1 \xi_2 \xi_3\ P_{10} \\
  \end{split}
\end{equation}

\noindent
where $ \xi_3 = 1 - \xi_1 - \xi_2 $. By naming $n_i$ the respective values
of the Bernstein basis functions linked to control point $P_i$ we can
shortly express \eqref{eq-cubicbez} as:

@@

\Figure{0.8\linewidth}{bezier}
{Cubic Bézier triangle with mesh of 10 control points.}

@@

The Bézier triangles are defined by a mesh of control points, however the
simulated object is usually described by a triangular mesh. We construct
the Bézier mesh by taking into account the curvature of the triangular
mesh. When constructing the mesh a special care has to be taken to maintain
the continuity across the nodes and edges. During the initialization phase
we partially employ the method described in \cite{Ubach2010} to maintain
$C^0$ continuity on the edges. Each of the control points on the edge is
computed as the intersection of:

\begin{enumerate}
    \item The plane perpendicular to the normal at the vertex.
    \item The plane that contains the curve of triangle's contour. The
        choice is arbitrary, but necessary to maintain $C^0$ continuity. 
        We choose the plane defined by the edge of the flat triangle and
        average of the two normals at vertices of the edge. 
    \item The plane perpendicular to the edge of the flat triangle placed at
        $1/3$ of the edge length.
\end{enumerate}

@@

That way if the element is not flat the central point is slightly elevated
and not in the plane of other nodes thus keeping the curvature of the
element.

@@

In figure \ref{fig-fps} we present performance of the element in terms of
frames per second. All tests were performed on machine equipped Dual-Core
AMD Opteron Processor 2218 and 3 GB of RAM and conjugate gradient solver
has been to solve the system. Our implementation is without optimizations
and isn't parallel, which means only single CPU core was used for the
simulation. The values reported are for the raw physical simulation with
and without the frame fixing method described in subsection
\ref{sec-corot}. We also present results including the mapping of
high-resolution mesh onto the mechanical mesh for visualisation purposes or
collisions.

\Figure{\linewidth}{fps}
{Performance in frames per seconds for different number of elements. Shows
values for element with frame fixing (B) method described in subsection
\ref{sec-corot} and without it (A), and with high-resolution mesh of 2048
(C) or 3200 (D) triangles mapped onto the mechanical mesh.}

@@

\subsection{Applications}

The presented element has been successfully implemented in SOFA framework
\cite{Faure2012} for real-time surgical simulations. It has been evaluated
on several applications.

% TODO: !!! When ANONYMISING do something with the citation !!!
One being a tool for pre-operative planning of corrective surgery of
congenital heart diseases in infants \cite{Kislinskiy2012}. The cardiac
surgery is a complicated process where several arterial vessels have to be
cut, patched and reconnected. The element has been used for to model
deformations of blood vessels after suturing and/or applying a artificial
patch. Figure \ref{fig-chd} presents a screenshots of the application
prototype.

\FigureW{0.75\linewidth}{chd}
{Screenshots of the simulation system prototype for different surgical
procedures used to repair a coarctation of an aortic arch. The screenshots
are enriched with overlay delineations.}

Another designed application is simulation of women reproductive organs
(vagaina/uterus/\ldots). 
The simulation is so as to evaluate
stress on a tissue during a birth.

@@

@@

On the other hand the choices provide a good trade-off between continuity
and simplicity. Only $ 3 \times 6~\mathrm{DOFs} = 18~\mathrm{DOFs}$ are
necessary while the Bézier triangle has theoretically $ 10 \times
3~\mathrm{DOFs} = 30~\mathrm{DOFs}$. For complex elements with high number
of nodes the interconnection between nodes increases, especially on the
vertices where the size of nodes involved depends also on number of
neighbouring elements. By keeping the element simple the sparsity of
stiffness matrix is not violated.

@@

    \item It is continuous on the edges and provides $C^1$ continuity in
    corners of the element. This is essential for good visual experience if
    high-resolution mesh is mapped on the mechanical mesh.

@@

Because of the formulation of Bézier triangle that is intrinsically present
there is a clean and obvious way of mapping high-resolution mesh for
better visualisation. We also provide a mapping of forces and velocities
between simulation mesh and mechanical mesh for collision detection/response
and interaction with the mesh which are essential component for the surgical
simulator.
